Experiences with Planning for Natural Language Generation

### 2009-06-04 trunk/Papers/ci-spark/experiments.tex
changes to section 4

We now return to the original question of the paper: is planning technology
ready for realistic applications in natural language generation? To investigate
this question we consider two sets of experiments, designed to evaluate the
performance of several planners on the NLG planning domains from the previous
section. Starting with the CRISP domain, we first present a scenario which
focuses on the generation of referring expressions with a tiny grammar
(Section~\ref{sec:exper-1:-sent}). We then look at a setting in which CRISP is
used for surface realization with the XTAG Grammar \citep{xtag01:_tr}, a
large-scale TAG grammar for English (Section~\ref{sec:exper-2:-sent-xtag}). In
the second set of experiments we investigate the GIVE domain. We begin with a
domain that is similar to the classic Gridworld
(Section~\ref{sec:exper-2:-minim}), and then add extra grid cells to the world
that are neither needed (Section~\ref{sec:experiment-3:-give}) nor even
accessible (Section~\ref{sec:experiment-4:-give}). We also investigate the role
that goal ordering plays in some of these problems. These experiments are
configured in a way that lets us explore the scalability of a planner's search
and preprocessing capabilities, and illustrates what we perceive to be one of
the main limitations of current off-the-shelf planners for our applications:
they often spend a long time computing ground instances, even when most of
these instances are not required during plan search.

@@

\begin{figure}[t]
  \centering
  \subfloat[position=bottom][(a) Minimal world]{%
      \includegraphics[height=2.5cm,clip=true,trim=0 -10px 0 -3px]{pic-buttons}}
  \qquad
  \subfloat[position=bottom][(b) World with extra grid cells]{%
      \includegraphics[height=2.5cm]{pic-empty-buttons}} \\
  %\subfloat[(c) Extra grid cells with inaccessible positions]{%
  %    \includegraphics[height=2.5cm]{pic-empty-inaccessible}}
  \caption{Experimental GIVE world configurations.}
  \label{fig:give-maps}
\end{figure}

\subsection{Experiment 3: Minimal GIVE worlds}
\label{sec:exper-2:-minim}

@@

Results for the $h=20$ case, with the grid width $N$ ranging from $1$ to $40$,
are shown in Figure~\ref{fig:give-minimal}. In the unordered case
(Figure~\ref{fig:give-minimal}(a)), the most obvious result is that some of the
planners tested---Metric-FF, FF(h$_a$), and C$^3$---are unable to solve any
problems beyond $N=24$ on our experimentation machine within the memory limit of
2 GB.\footnote{\todo{describe testing environment}} While FF, LAMA, and SGPLAN
are able to solve all problem instances up to $N=40$, the total running time
varies greatly between these planners. For instance, FF takes almost 35 seconds
to solve the $N=40$ problem, while LAMA takes around 6.5 seconds. SGPLAN shows
impressive performance, generating a 240 step plan in the $N=40$ case in well
less than a second. In the ordered case (Figure~\ref{fig:give-minimal}(b)), we
again have the situation where Metric-FF, FF(h$_a$), and C$^3$ are unable to
solve all problem instances. Furthermore, both SGPLAN and LAMA, which performed
well on the unordered problem, now perform much worse than FF: FF takes 39
seconds for the $N=40$ case, while SGPLAN takes 50 seconds and LAMA takes 90
seconds. In real NLG systems, where response time is essential, running times
over a few seconds are unacceptable.

Preprocessing (parsing, grounding, etc.) time generally plays less of a role in
GIVE, compared with the sentence generation domain, however, its effects still
contribute significantly to the overall running time of a number of planners.
Figure~\ref{fig:give-minimal-grounding} shows the grounding time for FF, LAMA,
and SGPLAN on the minimal GIVE problems, compared with the total running time.
In the unordered variant of the minimal GIVE domain
(Figure~\ref{fig:give-minimal-grounding}(a)), the grounding time in LAMA and
SGPLAN accounts for a significant fraction of the total runtime: SGPLAN spends
around 40\% of its total runtime on preprocessing; for LAMA, this number
rises to at least 80\% for our test problems. For FF, the preprocessing time is
much less important than the search time, especially for large problem
instances. In the ordered case (Figure~\ref{fig:give-minimal-grounding}(b)), the
actual time spent on preprocessing is essentially unchanged from the unordered
case, and search time dominates the total runtime for all three planners.
However, overall, FF is now much better at controlling the search, compared with
the other planners and its performance on the unordered variant of the problem.

@@

@@

\subsection{Experiment 4: GIVE worlds with extra positions}
\label{sec:experiment-3:-give}

In the third experiment, we vary the structure of the GIVE world in order
to judge the effect that universe size has on the planning problem in this
domain. Starting with the ordinary GIVE world described in Experiment~2, we
extend the world map by adding another $w$ by $h$ empty ``junk'' positions
to the right of the minimal world, as shown in Figure~\ref{fig:give-maps}(b).
These new positions are not actually required in any plan, but extend the
size of the state space and approximate the situation in the actual GIVE
domain where most grid positions are never used. We leave the initial state
and goal untouched. As before, we generate a GIVE world description and
then convert it into a planning problem in PDDL.

Results for the $h=20$, $n=5$ case with $w$ ranging from $1$ to $70$ are
shown in Figure~\ref{fig:give-extracells}. As in Experiment~2, FF again
runs out of memory, this time at $w=17$, while SGPLAN easily solves inputs
beyond $w=70$. As in Experiment~2, our implementation of GraphPlan is
unable to generate any of these plans. Unlike Experiment 2, however, both
FF and SGPLAN now spend a substantial proportion of their time on
grounding. In SGPLAN, this translates to a ``parsing time'' (which we
assume includes grounding) which grows from 180 ms to 21.7 seconds as $w$
grows from $1$ to $75$. The rest of the runtime, which also includes the
search time, only grows from 400 ms to 2.3 seconds. This difference is
particularly dramatic given that the actual optimal plan in each case is an
identical plan of about 100 steps (i.e., the one we would have found in
Experiment 2 for the $h=20$, $n=5$ case). The planning times for these
instances are also concerning since times over a couple seconds will
negatively affect the overall response time of the system, which must react
in real time to user actions.

%\begin{figure}[t]
%  \centering
%  \includegraphics[width=0.75\columnwidth]{graph-exp3}
%  %\includegraphics[width=1\columnwidth]{pic-runtime-empty-world}
%  \caption{Results on the GIVE domain with junk
%    positions for $h=20$ and $n=5$. The horizontal axis is $w$.
%    The vertical axis is the runtime in seconds.}
%\end{figure}

@@

\subsection{Experiment 5: GIVE worlds with inaccessible positions}
\label{sec:experiment-4:-give}

In the final experiment, we consider a variation on the GIVE worlds from
Experiment~3. As before, we begin with a base grid of $2n$ by $h$
positions, with buttons at positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i
\leq n$. We then generate a second grid of $w$ by $h$ positions and place
an additional button in this grid. Unlike Experiment~3, the second grid
does not extend the first grid but is ``disconnected'' so its positions are
inaccessible from the first grid, as shown in
Figure~\ref{fig:give-maps}(c). Since the geometry of the grid makes it
impossible to construct a plan for pressing all the buttons in the world,
we instead investigate the time it takes a planner to arrive at the
conclusion that the problem cannot be solved.

Results for the $h=20$, $n=5$ case with $w$ ranging from 1 to 50 are shown
in Figure~\ref{fig:give-runtime-nosoln}.\footnote{FF and SGPLAN do
 not display runtimes when they fail to construct a plan so an external timing
 program was used to generate these results.}
In this case, the runtimes for FF rise sharply around $w=28$, producing
times that are unacceptable for an NLG system in practice. FF is unable
to complete problem instances above $w=35$. SGPLAN, by comparison, performs
exceptionally well and is able to complete problems instances up to $w=50$
in about 0.3 seconds and $w=80$ in about 0.6 seconds. These results
demonstrate that SGPLAN has a significant advantage over FF in this
experiment. 
%Our implementation of GraphPlan is unable to complete any of
%these problem instances.

%\begin{figure}
%  \centering
%  \includegraphics[width=0.85\columnwidth]{graph-exp4}
%  %\includegraphics[width=1\columnwidth]{pic-runtime-empty-world}
%  \caption{Results on the GIVE domain with inaccessible
%    positions for $h=20$ and $n=5$. The horizontal axis is $w$.
%    The vertical axis is the runtime in seconds.}
%  \label{fig:give-runtime-nosoln}
%\end{figure}

>>>>

We now return to the original question of the paper: is planning technology
ready for realistic applications in natural language generation? To investigate
this question we consider two sets of experiments, designed to evaluate the
performance of several planners on the NLG planning domains from the previous
section. Starting with the CRISP domain, we first present a scenario which
focuses on the generation of referring expressions with a tiny grammar
(Section~\ref{sec:exper-1:-sent}). We then look at a setting in which CRISP is
used for surface realization with the XTAG Grammar \citep{xtag01:_tr}, a
large-scale TAG grammar for English (Section~\ref{sec:exper-2:-sent-xtag}). In
the second set of experiments we investigate the GIVE domain. We begin with a
domain that is similar to the classic Gridworld
(Section~\ref{sec:exper-3:-minim}), and then add extra grid cells to the world
that are not necessary to complete the task
(Section~\ref{sec:experiment-4:-give}). We also investigate the role that goal
ordering plays in these problems. These experiments are configured in a way that
lets us explore the scalability of a planner's search and preprocessing
capabilities, and illustrates what we perceive to be one of the main limitations
of current off-the-shelf planners for our applications: they often spend a long
time computing ground instances, even when most of these instances are not
required during plan search.

@@

\begin{figure}[t]
  \centering
  \subfloat[position=bottom][(a) Minimal GIVE world]{%
      \includegraphics[height=2.5cm,clip=true,trim=0 -10px 0 -3px]{pic-buttons}}
  \qquad
  \subfloat[position=bottom][(b) GIVE world with extra grid cells]{%
      \includegraphics[height=2.5cm]{pic-empty-buttons}} \\
  %\subfloat[(c) Extra grid cells with inaccessible positions]{%
  %    \includegraphics[height=2.5cm]{pic-empty-inaccessible}}
  \caption{Experimental GIVE world configurations.}
  \label{fig:give-maps}
\end{figure}

\subsection{Experiment 3: Minimal GIVE worlds}
\label{sec:exper-3:-minim}

@@

@@

Results for the $h=20$ case, with the grid width $N$ ranging from $1$ to $40$,
are shown in Figure~\ref{fig:give-minimal}. In the unordered case
(Figure~\ref{fig:give-minimal}(a)), the most obvious result is that some of the
planners tested---Metric-FF, FF(h$_a$), and C$^3$---are unable to solve any
problems beyond $N=24$ on our experimentation machine within the memory limit of
2 GB.\footnote{All runtimes in Sections~\ref{sec:exper-3:-minim} and
 \ref{sec:experiment-4:-give} were measured on a single core of an Intel Xeon
 CPU running at 3GHz, under Linux. All runtimes are averaged over three runs of
 the planners. Only 32-bit versions of the planners were used for testing in
 each case.}
While FF, LAMA, and SGPLAN are able to solve all problem instances up to $N=40$,
the total running time varies greatly between these planners. For instance, FF
takes almost 35 seconds to solve the $N=40$ problem, while LAMA takes around 6.5
seconds. SGPLAN shows impressive performance on $N=40$, generating a 240 step
plan in well under a second. In the ordered case
(Figure~\ref{fig:give-minimal}(b)), we again have the situation where Metric-FF,
FF(h$_a$), and C$^3$ are unable to solve all problem instances. Furthermore,
both SGPLAN and LAMA, which performed well on the unordered problem, now perform
much worse than FF: FF takes 39 seconds for the $N=40$ case, while SGPLAN takes
50 seconds and LAMA takes 90 seconds. In real NLG systems, where response time
is essential, running times over a few seconds are unacceptable.

Preprocessing (parsing, grounding, etc.) time generally plays less of a role in
GIVE, compared with the sentence generation domain, however, its effects still
contribute significantly to the overall running time of a number of planners.
Figure~\ref{fig:give-minimal-grounding} shows the grounding time for FF, LAMA,
and SGPLAN on the minimal GIVE problems, compared with the total running time.
In the unordered variant of the minimal GIVE domain
(Figure~\ref{fig:give-minimal-grounding}(a)), the grounding time in LAMA and
SGPLAN accounts for a significant fraction of the total runtime: SGPLAN spends
around 40\% of its total runtime on preprocessing; for LAMA, this number
rises to at least 80\% for our test problems. For FF, the preprocessing time is
much less important than the search time, especially for large problem
instances. In the ordered case (Figure~\ref{fig:give-minimal-grounding}(b)), the
actual time spent on preprocessing is essentially unchanged from the unordered
case, and search time dominates the total runtime for all three planners.
Overall, however, FF is now much better at controlling the search, compared with
the other planners and its performance on the unordered variant of the problem.

@@

\subsection{Experiment 4: GIVE worlds with extra grid cells}
\label{sec:experiment-4:-give}

In our last set of experiments, we vary the structure of the GIVE world in order
to judge the effect that universe size has on the resulting planning problem.
Starting with the GIVE world described in Experiment~3, we extend the world map
by adding another $w$ by $h$ empty cell positions to the right of the minimal
world, as shown in Figure~\ref{fig:give-maps}(b). These new positions are not
actually required in any plan, but extend the size of the state space and
approximate the situation in the actual GIVE domain where most grid positions
are never used. We leave the initial state and goal untouched and, again,
consider both unordered and ordered variants of the problem.

Results for the $h=20$, $n=10$ case with $w$ ranging from $1$ to $40$ are shown
in Figure~\ref{fig:give-extracells}. As in Experiment~3, a number of planners
again fail to solve all the problems: Metric-FF, FF(h$_a$), and C$^3$ solve only
a few instances, while FF only scales to $w=23$. In the unordered version of the
domain, SGPLAN easily solves inputs beyond $w=40$ in well less than a second.
LAMA is also reasonably successful on these problems, however, its runtimes grow
more quickly than SGPLAN, with LAMA taking almost 5 seconds to solve the $w=40$
problem instance. In the ordered case, we again see behaviour similar to that of
Experiment~3: for the problem instances FF is able to solve, it performs
significantly better than LAMA and SGPLAN. (SGPLAN's long term runtime
appears to be growing at a slower rate than FF's, and so even if FF could be
scaled to larger problem instances, it seems possible that SGPLAN might overtake
FF as the better performer.) However, the overall planning times for most of
these instances are concerning since times over a couple seconds will negatively
affect the response time of an NLG system, which must react in real time to user
actions.

Finally, we also performed a set of experiments designed to investigate the
tradeoff between grounding time and search time on certain grid configurations.
For these experiments, we initially fixed the size of the grid and then varied
the number of buttons $b$ in the world, thereby creating a series of
``snapshots'' of particular extra-cell GIVE domains.
Figure~\ref{fig:give-buttons} shows the results of these experiments for the FF
and SGPLAN planners, for a fixed size grid of height 20 and width 40, and the
number of buttons $b$ ranging from 1 to 40. In each case, the amount of time a
planner spends on grounding is relatively unchanged as we vary the number of
buttons in a grid, while the search time continues to rise (sometimes quite
dramatically), as $b$ increases (we saw a similar effect for other grid
configurations we tried). This observation has important consequences for the
design of our GRID worlds: changing the underlying domain structure, even
minimally, may result in significant---and often unexpected---performance
differences for the planners that must operate in these domains.

@@

%In the final experiment, we consider a variation on the GIVE worlds from
%Experiment~3. As before, we begin with a base grid of $2n$ by $h$
%positions, with buttons at positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i
%\leq n$. We then generate a second grid of $w$ by $h$ positions and place
%an additional button in this grid. Unlike Experiment~3, the second grid
%does not extend the first grid but is ``disconnected'' so its positions are
%inaccessible from the first grid, as shown in
%Figure~\ref{fig:give-maps}(c). Since the geometry of the grid makes it
%impossible to construct a plan for pressing all the buttons in the world,
%we instead investigate the time it takes a planner to arrive at the
%conclusion that the problem cannot be solved.

%Results for the $h=20$, $n=5$ case with $w$ ranging from 1 to 50 are shown
%in Figure.\footnote{FF and SGPLAN do
% not display runtimes when they fail to construct a plan so an external timing
% program was used to generate these results.}
%In this case, the runtimes for FF rise sharply around $w=28$, producing
%times that are unacceptable for an NLG system in practice. FF is unable
%to complete problem instances above $w=35$. SGPLAN, by comparison, performs
%exceptionally well and is able to complete problems instances up to $w=50$
%in about 0.3 seconds and $w=80$ in about 0.6 seconds. These results
%demonstrate that SGPLAN has a significant advantage over FF in this
%experiment.
